---
title: "Model Selection with Train/Validate/Test"
format: html
editor: visual
---

# Assignment 3: Model Selection with Train/Validate/Test

You’ll turn this `.qmd` file in as your final modeling assignment. Save, commit, and push to GitHub. Then, go to Canvas and type "Submitted" under the assignment submission. Assignment due Sunday 4/6 at 11:59 pm.

------------------------------------------------------------------------

In this activity, you'll practice selecting the best predictive model using a **train/validate/test split**. This is one step beyond a train/test split.

You’ll compare multiple models using both *in-sample evaluation* (like AIC and ANOVA) and *out-of-sample validation* (using RMSE). You'll then evaluate your final model on a held-out test set.

## Why Use Train/Validate/Test Instead of Just Train/Test?

In a basic train/test split:

-   You train your model on one part of the data

-   You test its performance on the rest

But what if you want to compare multiple models?

If you use the test set to pick the best one, you've “peeked” — and the test set is no longer a fair estimate of how your model performs on truly unseen data.

So we add a **validation set**:

-   **Training set** → Fit multiple models

-   **Validation set** → Choose the best model

-   **Test set** → Evaluate final model performance

This approach helps prevent overfitting and gives you a more realistic estimate of how your model will perform in the real world.

## Set Up Packages

Add packages as needed.

```{r}
# Setup
library(tidyverse)
library(caret)
library(Metrics)
set.seed(42)  # for reproducibility
```

## Dataset Requirements

You may choose your own dataset for this assignment.

Choose a dataset that:

-   Has a numeric outcome variable you want to predict

-   Contains at least 3-4 predictors (numeric or categorical)

-   Is either:

    -   A built-in dataset in R (e.g., `diamonds`, `Boston`, `iris`, `mtcars`, `airquality`, `penguins`, etc.)
    -   From your final project
    -   Any other dataset we've used in class

> If you're not sure what dataset to use, try `Boston`:

```{r}
library(MASS)
data <- Boston # read in your data here
```

## Step 1: Split the Data

Split the data into: 60% training, 20% validation, and 20% test

```{r}
# edit below as needed
train_index <- createDataPartition(data$medv, p = 0.6, list = FALSE)
train_data <- data[train_index, ] # training data
temp_data <- data[-train_index, ]

val_index <- createDataPartition(temp_data$medv, p = 0.5, list = FALSE)
val_data <- temp_data[val_index, ] # validation data
test_data <- temp_data[-val_index, ] # test data
```

## Step 2: Fit Multiple Models

Create at least three models of increasing complexity:

```{r}
# edit below as needed
model_1 <- lm(medv ~ lstat, data = train_data)
model_2 <- lm(medv ~ lstat + rm + age, data = train_data)
model_3 <- lm(medv ~ lstat * rm + tax + ptratio, data = train_data)

summary(model_1)
summary(model_2)
summary(model_3)
```

**Questions:**

-   Which model seems to be the best fit according to the Adjusted R^2^ value?

> You may look at R² and Adjusted R² on the training set to help understand model fit, but to ultimately choose the best model, you'll use RMSE on the validation set below.\
> RMSE gives you a more honest view of how well your model predicts on new data.

Model_3 appears to be the best at predicting, this is because it has the lowest R scores.

## Step 3: Compare Using AIC and ANOVA

> -   AIC helps you compare model fit while penalizing complexity
>
> -   ANOVA tests whether adding predictors significantly improves the model

```{r}
# AIC
AIC(model_1)
AIC(model_2)
AIC(model_3)

# ANOVA for nested comparisons
anova(model_1, model_2)
anova(model_2, model_3)
```

**Questions:**

-   Which model has the lowest AIC?

    model 3

-   Are the improvements in fit (from the `anova` output) statistically significant?

    Yes, they are very statistically significant.

## Step 4: Evaluate on the Validation Set (RMSE) (new)

The validation set allows us to compare models fairly and reevaluate our choices before making a final decision. If a model performs well on training but poorly on validation, we might consider simplifying or adjusting the model before moving on to the test set.

```{r}
# edit below as needed
rmse(val_data$medv, predict(model_1, val_data))
rmse(val_data$medv, predict(model_2, val_data))
rmse(val_data$medv, predict(model_3, val_data))
```

```{r}
rmse(train_data$medv, predict(model_1, train_data))
rmse(train_data$medv, predict(model_2, train_data))
rmse(train_data$medv, predict(model_3, train_data))
```

**Questions:**

-   Which model performed best on the validation set?

    model 3

-   Does that match what AIC/ANOVA suggested?

    Yes, it matches. model 3 has the lowest score by a pretty decent amount, and appears to be much more accurate in predictions.

## Step 5: Choose the Best Model

Pick the model with the best validation RMSE. Assign it to a variable called `final_model`. This isn't a "required" step, but it keeps things neat when you only need to define the final model in one spot.

```{r}
final_model <- model_3
```

## Step 6: Test the Final Model

Now evaluate your chosen model on the test set:

```{r}
# edit below as needed
rmse(test_data$medv, predict(final_model, test_data))
```

```{r}
rmse(test_data$medv, predict(model_1, test_data))
rmse(test_data$medv, predict(model_2, test_data))
rmse(test_data$medv, predict(model_3, test_data))
```

**Questions:**

-   Is the test RMSE close to the validation RMSE?

    not really- it is significantly lower than the validation data

-   What does that say about how well the model generalizes?

    It generalized quite well. It performed better on the test data than it did on the validation data. Because the valid/test split was pretty small, the validation data was likely harder to predict just by chance.

## Step 7: Compare All RMSE Values

```{r}
# edit below as needed
rmse(train_data$medv, predict(final_model, train_data)) # training set
rmse(val_data$medv, predict(final_model, val_data)) # validation set
rmse(test_data$medv, predict(final_model, test_data)) # test set
```

**Questions:**

-   Is there a big gap between training and validation/test RMSE? If so, does that suggest overfitting?

    yes, there is a significant gap. However, this does not necessarily mean there is overfitting.

## Summary Questions

Answer the following. Use full sentences.

1.  Which model did you choose, and why?

    I chose model 3 because it had the most significant capabilities in predicting the training data

2.  What were the AIC values for each model?

    ```         
    [1] 2035.642
    [1] 1964.25
    [1] 1836.726
    ```

3.  Did ANOVA support the improvements?

    Yes

4.  What were the RMSE values for training, validation, and test sets?

    train:

    ```         
    [1] 6.669002
    [1] 5.896033
    [1] 4.755834
    ```

    validation:

    ```         
    [1] 5.659532
    [1] 5.407137
    [1] 4.788403
    ```

    test:

    ```         
    [1] 5.176061
    [1] 4.490713
    [1] 3.814994
    ```

5.  How confident are you that your model will generalize well?

    I am pretty confident that my model can generalize well, however I am concerned there is overfitting. Model 3 likely performs well, however i think more analysis is needed to determine if there is overfitting.

*Reminder: Your chosen model should balance good in-sample fit (R², AIC) with strong out-of-sample performance (validation RMSE), and generalize well to the test set. You don’t have to pick the “most complex” model — just the one that performs reliably and addresses the research question.*
